---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "80%",
  dpi = 150,
  fig.width = 9,
  fig.height = 3.2
)
```

# bafratio

<!-- badges: start -->
[![R-CMD-check](https://github.com/zengzhechun/bafratio/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/zengzhechun/bafratio/actions/workflows/R-CMD-check.yaml)
[![License: MIT](https://img.shields.io/badge/license-MIT-blue.svg)](https://opensource.org/licenses/MIT)
<!-- badges: end -->

**Quantify how much of your "effect" is actually bias.**

In observational studies, a statistically significant association is not
necessarily a causal one. Empirical calibration with negative control
outcomes (Schuemie et al., 2014) tells you *whether* residual systematic
bias plausibly explains your estimate, via a calibrated p-value. But a
p-value is a binary verdict: it mixes the *size* of the bias with the
*precision* of the estimate, and it says nothing about the magnitude of
bias relative to the effect that survives calibration.

`bafratio` separates the two with a pair of equivalent metrics. The
primary metric is the **bias attribution fraction (BAF)**, the share of the total
calibrated signal attributable to systematic bias, on a bounded 0-1 scale:

$$
\mathrm{BAF} = \frac{|\mu_B|}{|\mu_B| + |\log RR_{\text{calibrated}}|}
= \frac{|\text{systematic bias}|}{|\text{bias}| + |\text{effect remaining after calibration}|}
$$

The auxiliary **bias-effect ratio (BER)** expresses the same information as
an unbounded ratio: $\mathrm{BER} = \mathrm{BAF} / (1 - \mathrm{BAF})$.

* **BAF > 0.5** (BER > 1) — bias accounts for more than half of the
  calibrated signal (*bias-dominated*). The "effect" is more likely an
  artifact than a fact.
* **1/3 ≤ BAF ≤ 0.5** (0.5 ≤ BER ≤ 1) — bias and signal are comparable
  (*mixed*). Interpret with caution.
* **BAF < 1/3** (BER < 0.5) — the calibrated signal clearly exceeds the
  bias (*effect-dominated*). The association survives the bias audit.

`bafratio` computes BAF and BER with logit-scale (equivalently log-scale)
bootstrap confidence intervals, bootstrap-median point estimates that
cannot fall outside their own CI, a Fieller confidence set for the
underlying ratio, three-zone classification, leave-one-out sensitivity
analysis over the negative controls, and publication-ready `ggplot2`
visualizations designed for clinical audiences.

## Installation

```{r, eval = FALSE}
# From GitHub (development version)
# install.packages("remotes")
remotes::install_github("zengzhechun/bafratio")
```

## Quick start

The package ships a fully synthetic dataset, `sim_nc` / `sim_est`, that
reproduces a classic *healthy-adherer* scenario: a drug that appears to
protect against an outcome (RR = 0.83) purely because of systematic error.
No real patient data are involved.

```{r quickstart}
library(bafratio)

fit <- baf_analyze(
  logRr    = sim_est$logRr,   # primary estimate: log RR = -0.184 (RR = 0.83)
  seLogRr  = sim_est$seLogRr,
  ncLogRr  = sim_nc$logRr,    # 12 negative control outcomes
  ncSeLogRr = sim_nc$seLogRr,
  ncNames  = sim_nc$outcome,
  nBoot = 2000, seed = 42,
  loo = TRUE
)

fit
```

```{r gauge, fig.height = 2.6}
plot_baf_gauge(fit)
```

The uncalibrated estimate looks protective (RR = 0.83). Twelve negative
controls reveal a systematic protective bias ($\mu_B \approx -0.21$).
After calibration the effect vanishes (calibrated RR = 1.02, calibrated
p = 0.65), and the bias attribution fraction is BAF = 0.88 (95% CI, 0.75-0.99):
bias accounts for roughly nine tenths of the calibrated signal. Read BAF
near 1 as *signal saturation by bias*, not as a precise fraction.

```{r nullplot, fig.height = 4.2}
plot_null(fit)
```

The diamond (primary estimate) sits inside the cloud of negative controls:
the visual signature of a bias-dominated association.

```{r calibration, fig.height = 2.9}
plot_calibration(fit)
```

```{r loo, fig.height = 4.2}
plot_loo(fit$loo)
```

## Two-layer screening

The full workflow above tells you *how big* the bias is. When you must
decide, for many exposure-outcome pairs at once, whether each estimate can
serve as usable **effect evidence**, use `baf_screen()` (or its BER-scale
alias `bf_screen()`). It applies a two-layer rule:

1. **Layer 1 (signal existence).** The calibrated p-value must be below
   `cal_p_threshold` (default 0.05). Otherwise the verdict is
   `"insufficient-evidence"` and Layer 2 is not entered.
2. **Layer 2 (bias share).** Conditional on a real signal, `baf_screen()`
   looks up, in the reference simulation calibration `baf_reliability`, the
   probability `P(bias-dominated)` that the *true* regime is bias-dominated,
   given the BAF point estimate and the *width* of its 95% CI (a wide CI means
   the BAF is poorly pinned down, so the conservative maximum of the
   narrow-CI and wide-CI probabilities is reported). The verdict then follows
   fixed bands: `< 0.15` effect-evidence, `0.15-0.45` mixed /
   hypothesis-generating, `0.45-0.65` competitive / inconclusive,
   `>= 0.65` not-effect-evidence. The `explain()` method returns a
   plain-English justification.

```{r screen, fig.height = 3.2}
sc <- baf_screen(
  logRr     = sim_est$logRr,
  seLogRr   = sim_est$seLogRr,
  ncLogRr   = sim_nc$logRr,
  ncSeLogRr = sim_nc$seLogRr,
  nBoot = 500, seed = 42
)
sc
explain(sc)

plot(sc)            # BAF gauge annotated with the screening verdict
```

The empirical null and the BAF 95% CI come from the same fit as
`baf_analyze()`; `baf_screen()` merely adds the Layer-1/Layer-2 decision and
the reliability lookup. For a single pair you can pass a `bafratio` object
directly (`baf_screen(fit)`), skipping the recomputation. The
`baf_reliability` table (12 BAF bins, 960,000 simulated repetitions) is shipped
with the package and is the only data `baf_screen()` needs beyond the fit.

## What the metrics mean, in plain terms

| Question | Metric | Scale | Read as |
|---|---|---|---|
| Of the signal that survives calibration, how much is bias? | **BAF** | 0-1 | bias-dominated > 0.5; mixed 1/3-0.5; effect-dominated < 1/3 |
| How many times larger is the bias than the residual effect? | **BER** | 0-∞ | BAF/(1-BAF); same three zones at 1 and 0.5 |
| Of the *uncalibrated* association, how much is bias? | OBF | 0-∞ | $\|\mu_B\| / \|\log RR_{\text{obs}}\|$; descriptive companion |

Report BAF (with its CI and zone) alongside the calibrated p-value and the
calibrated RR. The calibrated p-value answers *is there evidence of an
effect after accounting for bias?* BAF answers *how much of the surviving
signal is bias?* Neither is a substitute for the other.

## Main functions

| Function | Purpose |
|---|---|
| `baf_estimate()` | fit the empirical null; BAF and BER point estimates, calibrated RR and p |
| `baf_bootstrap()` | bootstrap CIs (log/logit scale) + bootstrap-median point estimates |
| `ber_classify()`, `baf_classify()` | three-zone classification (ratio / fraction scale) |
| `baf_fieller()` | Fieller confidence set for the ratio $\mu_B / \tilde\psi$ |
| `baf_loo()` | leave-one-out sensitivity over the negative controls |
| `baf_diagnostics()` | standardized residuals, Shapiro-Wilk, Q-Q data |
| `baf_analyze()` | the whole workflow in one call |
| `baf_screen()`, `bf_screen()` | two-layer screening (signal-existence + bias-share) into an effect-evidence verdict |
| `baf_rules()` | evaluate the companion paper's reporting rules `R0`-`R4`, `C1`, `C2` from a BAF, its interval and the calibrated p |
| `explain()` | plain-English justification of a `baf_screen` verdict |
| `baf_reliability` | shipped reference calibration table (12 bins, 960,000 reps) used by Layer 2 |
| `plot_baf_gauge()`, `plot_gauge()`, `plot_null()`, `plot_calibration()`, `plot_qq()`, `plot_loo()`, `plot_boot()` | visualizations |

## Background and assumptions

The approach rests on the OHDSI empirical calibration framework
(Schuemie et al., 2014, 2018) and on **negative control outcomes**:
outcomes that share the confounding structure of the primary endpoint but
cannot plausibly be caused by the exposure (Lipsitch et al., 2010). The key
assumption is **bias exchangeability**: the residual bias affecting the
primary estimate is drawn from the same distribution as the biases visible
in the negative controls. Like the no-unmeasured-confounding assumption it
diagnoses, bias exchangeability cannot be verified from the data; its
plausibility must be argued from domain knowledge, and the leave-one-out
and Fieller analyses probe how far the conclusions depend on it.

Do not use the metrics when negative controls are few (< 5, ideally ≥ 10),
non-exchangeable, or point in opposite directions, or when the calibrated
effect is near zero (BAF near 1 signals saturation; the exact value should
not be over-interpreted).

## Companion paper

Zeng Z, Wang J, Zuo H, Shu L. Quantifying Systematic Bias in Observational
Causal Estimates Using Negative Controls: The Bias Attribution Fraction. Manuscript
v34 (2026). The simulation study (336 conditions, 200 repetitions each) and
the MIMIC-IV case study described there motivate the defaults used here:
BAF thresholds 0.5 and 1/3, logit-scale bootstrap with bootstrap-median
point estimates, and the conservative CI-based classification rule.

## References

Schuemie MJ, Ryan PB, DuMouchel W, Suchard MA, Madigan D. Interpreting
observational studies: why empirical calibration is needed to correct
p-values. *Statistics in Medicine* 2014;33(2):209-218.

Schuemie MJ, Hripcsak G, Ryan PB, Madigan D, Suchard MA. Empirical
confidence interval calibration for population-level effect estimation
studies in observational healthcare data. *PNAS* 2018;115(11):2571-2577.

Lipsitch M, Tchetgen Tchetgen E, Cohen T. Negative controls: a tool for
detecting confounding and bias in observational studies. *Epidemiology*
2010;21(3):383-388.
