---
title: "Auditing observational effects with the bias-effect ratio"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Auditing observational effects with the bias-effect ratio}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8.5,
  fig.height = 3.2,
  dpi = 130,
  out.width = "95%"
)
library(biasratio)
```

## The question this package answers

A significant p-value in an observational study answers only one question:
*if there were truly no association and no systematic error, how surprising
would this estimate be?* It says nothing about the second half of that
conditional. Residual confounding, selection bias, and measurement error
are not in the null hypothesis, so they are not in the p-value either.

Empirical calibration (Schuemie et al., 2014; OHDSI `EmpiricalCalibration`
package) repairs this by estimating the *systematic* error distribution
from **negative control outcomes** — outcomes the exposure cannot
plausibly cause — and re-calibrating the p-value against it. But the
output is still a p-value: a verdict, not a measurement.

The **bias-effect ratio** turns the verdict into a quantity. If the
empirical null has mean $\mu_B$ (the systematic bias, on the log-RR scale)
and the calibrated log effect is
$\log RR_{\mathrm{cal}} = \log RR_{\mathrm{obs}} - \mu_B$, then

$$
\mathrm{BER} = \frac{|\mu_B|}{|\log RR_{\mathrm{cal}}|}
$$

reads as *"the bias is X times as large as the effect that survives
calibration"*. The estimate is classified into three zones:

| Zone | BER | Reading |
| --- | --- | --- |
| bias-dominated | > 1 (or CI entirely above 1) | Bias exceeds the calibrated signal |
| competitive | 0.5-1 | Bias and signal are comparable |
| effect-dominated | < 0.5 (or CI entirely below 0.5) | Signal clearly exceeds bias |

## Case 1: a "protective effect" made of bias (simulated)

`sim_nc` and `sim_est` are fully synthetic data reproducing a
healthy-adherer scenario: 12 negative control outcomes plus one primary
estimate that looks protective (RR = 0.83, p < 0.001).

```{r case1}
fit <- ber_analyze(
  logRr     = sim_est$logRr,
  seLogRr   = sim_est$seLogRr,
  ncLogRr   = sim_nc$logRr,
  ncSeLogRr = sim_nc$seLogRr,
  ncNames   = sim_nc$outcome,
  nBoot     = 2000, seed = 42,
  loo       = TRUE, looBoot = 500
)
fit
```

Reading the printout top to bottom:

* The empirical null is centered at $\mu_B = -0.21$: every negative
  control leans "protective", so the study design itself manufactures
  protective-looking associations.
* Calibration subtracts that bias: RR moves from 0.83 to 1.02 and the
  calibrated p-value is 0.65.
* **BER = 8.7**: the bias is nearly nine times the residual signal.
  Classification: *bias-dominated*.

### The five-figure evidence board

```{r case1-gauge}
plot_gauge(fit)
```

The gauge is the one-look summary: the point and its bootstrap interval
sit far inside the red zone.

```{r case1-null, fig.height = 5.5}
plot_null(fit)
```

The fingerprint plot shows *why*: the diamond (primary estimate) lies
squarely on the bias ridge traced by the negative controls. An effect
estimate indistinguishable from known-zero effects is the visual
definition of bias dominance.

```{r case1-calibration, fig.height = 3.4}
plot_calibration(fit)
```

```{r case1-qq, fig.height = 4.2}
plot_qq(fit)
```

The Q-Q plot checks whether the negative controls behave like a normal
sample around the fitted null. Points near the diagonal mean the
calibration machinery is on solid ground; systematic curvature would
undermine the BER itself.

```{r case1-loo, fig.height = 4.6}
plot_loo(fit)
```

Leave-one-out analysis refits the entire pipeline excluding each negative
control in turn. The classification never flips: no single control is
carrying the conclusion.

## Case 2: the same audit on real data (sccs)

The `sccs` dataset shipped with `EmpiricalCalibration` contains
self-controlled case series estimates for sertraline and upper GI
bleeding, with 45 negative control drugs (Schuemie et al., 2014).

```{r case2}
data(sccs, package = "EmpiricalCalibration")
nc <- sccs[sccs$groundTruth == 0, ]
pc <- sccs[sccs$groundTruth == 1, ]

fitSccs <- ber_analyze(pc$logRr, pc$seLogRr, nc$logRr, nc$seLogRr,
                       ncNames = nc$drugName,
                       nBoot = 2000, seed = 42)
fitSccs
```

BER = 13.3 [5.4, 240.3]: despite 45 controls and a highly significant
raw estimate, the residual bias is an order of magnitude larger than the
calibrated signal. The lower confidence bound alone exceeds 5 — the
bias-dominated verdict is itself statistically secure.

```{r case2-gauge}
plot_gauge(fitSccs)
```

## Case 3: an effect that survives the audit (cohortMethod)

The `cohortMethod` dataset (isoniazid and acute liver injury, 30 negative
controls) is the mirror image: a strong, well-known true effect.

```{r case3}
data(cohortMethod, package = "EmpiricalCalibration")
nc2 <- cohortMethod[cohortMethod$groundTruth == 0, ]
pc2 <- cohortMethod[cohortMethod$groundTruth == 1, ]

fitCm <- ber_analyze(pc2$logRr, pc2$seLogRr, nc2$logRr, nc2$seLogRr,
                     ncNames = nc2$drugName,
                     nBoot = 2000, seed = 42)
fitCm
```

BER = 0.03 [0.00, 0.18]: the calibrated effect (RR = 4.2) dwarfs the
estimated bias. The entire bootstrap interval lies below 0.5, so the
classification is *effect-dominated* — exactly what a known hepatotoxin
should produce, and a useful positive control for the method itself.

```{r case3-gauge}
plot_gauge(fitCm)
```

## Workflow anatomy

`ber_analyze()` is a convenience wrapper. Every stage is available
individually, which matters for pipelines and for auditing the audit:

```{r anatomy}
est  <- ber_estimate(sim_est$logRr, sim_est$seLogRr,
                     sim_nc$logRr, sim_nc$seLogRr)   # point estimate
boot <- ber_bootstrap(sim_est$logRr, sim_est$seLogRr,
                      sim_nc$logRr, sim_nc$seLogRr,
                      nBoot = 2000, seed = 42)         # log-scale CI
cls  <- ber_classify(est$ber, boot$ci_lo, boot$ci_hi)  # three-zone verdict
diag <- ber_diagnostics(sim_nc$logRr, sim_nc$seLogRr)  # null adequacy
loo  <- ber_loo(sim_est$logRr, sim_est$seLogRr,
                sim_nc$logRr, sim_nc$seLogRr,
                ncNames = sim_nc$outcome, nBoot = 500, seed = 42)

c(ber = est$ber, ci_lo = boot$ci_lo, ci_hi = boot$ci_hi)
cls
diag$shapiro_p   # normality screen on standardized residuals
```

Bootstrap intervals are computed as percentile intervals **on the log
scale** and exponentiated. BER is a strictly positive, right-skewed ratio
estimator; the log transform yields an approximately symmetric bootstrap
distribution and intervals with sane order-of-magnitude coverage.

## Guardrails

* **Garbage controls in, garbage BER out.** Negative controls must share
  the bias mechanism of the primary estimate. The BER formalizes the
  calibration; it cannot rescue a broken control set.
* **BER is a diagnostic, not a decision rule.** "Bias-dominated" means
  the data cannot separate effect from bias — not that no effect exists.
* **Mind K.** With fewer than about 10 negative controls the empirical
  null is fragile; always run `ber_loo()` and look at `plot_qq()`.
* **Report the interval.** A BER of 1.2 [0.9, 40] and 1.2 [1.1, 1.3] are
  the same point and opposite stories.

## References

* Schuemie MJ, Ryan PB, Dumouchel W, Suchard MA, Madigan D. Interpreting
  observational studies: why empirical calibration is needed to correct
  p-values. *Statistics in Medicine* 33(2):209-218, 2014.
* Schuemie MJ, Hripcsak G, Ryan PB, Madigan D, Suchard MA. Empirical
  confidence interval calibration for population-level effect estimation
  studies in observational healthcare data. *PNAS* 115(11):2571-2577,
  2018.

```{r}
sessionInfo()
```
