---
title: "Quick start: auditing an effect with the bias attribution fraction"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Quick start: auditing an effect with the bias attribution fraction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8.5,
  fig.height = 3.4,
  dpi = 130,
  out.width = "95%"
)
library(biasratio)
```

## In ten minutes

This vignette walks through the whole `biasratio` workflow on the package's
built-in synthetic data, `sim_nc` (12 negative control outcomes) and
`sim_est` (one primary estimate). The scenario is a classic *healthy-adherer*
bias: a drug looks protective (RR = 0.83) purely because of systematic
error. No real patient data are involved.

Load the data and fit the full analysis in one call:

```{r analyze}
fit <- ber_analyze(
  logRr     = sim_est$logRr,
  seLogRr   = sim_est$seLogRr,
  ncLogRr   = sim_nc$logRr,
  ncSeLogRr = sim_nc$seLogRr,
  ncNames   = sim_nc$outcome,
  nBoot     = 2000,
  seed      = 42,   # optional; reproducibility is your responsibility
  loo       = TRUE
)
fit
```

## Read the result

* **BAF = 0.88** — about nine tenths of the calibrated signal is systematic
  bias. BAF > 0.5 means *bias-dominated*: the "effect" is more likely an
  artifact than a fact.
* The calibrated RR (= 1.02) and calibrated p (= 0.65) already tell you the
  naive association vanishes after calibration; BAF tells you *how much* of
  the surviving signal is bias.
* The CI-based classification (using the bootstrap interval) agrees:
  *bias-dominated*.

## Plots

`plot(fit)` defaults to the bounded **BAF gauge** (it lives on [0, 1], so a
wide bootstrap CI can never squeeze the zones):

```{r gauge, fig.width = 7, fig.height = 2.6}
plot(fit)
```

For the ratio-scale view, use `plot_gauge(fit)`. The empirical null, the
calibration, and the diagnostics are also one call each:

```{r others, fig.width = 8.5, fig.height = 3}
plot_null(fit)
plot_calibration(fit)
```

## Small-sample robustness

With only 12 controls, the plug-in calibrated p-value is materially
anti-conservative (about 11% Type I error at nominal 5%, per the companion
paper's eAppendix 13). Pass `method = "robust"` to apply the variance
inflation plus t(K−2) correction:

```{r robust}
est <- ber_estimate(sim_est$logRr, sim_est$seLogRr,
                    sim_nc$logRr, sim_nc$seLogRr, method = "robust")
est$cal_p_plugin
est$cal_p_robust
```

Here the conclusion is unchanged (both p-values are far from 0.05), but for
studies with fewer than ~20 controls the robust value is the more honest
one and should be reported as a sensitivity.

## Fieller set and leave-one-out

The Fieller confidence set for the underlying ratio degrades honestly when
the calibrated effect is near zero. With a fitted object you no longer have
to dig `var(boot$mu_draws)` out by hand:

```{r fieller}
boot <- fit$bootstrap
fieller(fit, boot)
```

Leave-one-out sensitivity (already computed above with `loo = TRUE`) shows
whether a single control drives the conclusion:

```{r loo, fig.width = 8.5, fig.height = 3}
plot_loo(fit$loo)
```

All 12 exclusions remain bias-dominated, so the conclusion is not owed to a
single influential negative control.
